---
title: "A Century of World Cup Football"
author: "Niklas Grau (ngrau3035@floridapoly.edu)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    fig_width: 8
    fig_height: 5
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(tidyverse)
library(scales)
```

# Introduction

This little project looks at every match played at the FIFA World Cup
from the first tournament in Uruguay in 1930 up to Brazil in 2014. The
file `WorldCupMatches.csv` lists one row per game and records who
played, how many goals each side scored, where the game took place, and
how many people came to watch.

I grew up watching the World Cup with my family, so when I saw this
dataset in the list I knew right away it was the one I wanted to dig
into. My goal was not to build something fancy. I wanted three clean
charts that each answer one simple question a casual fan might actually
ask.

# Loading the data

I keep the raw download in the `data/` folder and read it from there so
the whole report runs the same way on any machine.

```{r load}
matches_raw <- read_csv("../data/WorldCupMatches.csv", show_col_types = FALSE)

dim(matches_raw)
```

# Getting the data ready

The first thing I noticed is that the raw file is messy. It has
thousands of completely blank rows hanging off the bottom, and many of
the real games are repeated. Before doing anything else I drop the empty
rows and remove exact duplicates.

```{r clean1}
matches <- matches_raw %>%
  filter(!is.na(MatchID)) %>%   # blank rows have no match id
  distinct()                    # drop repeated rows

dim(matches)
```

That takes the file from over four thousand rows down to the real set of
games.

A second problem hides inside the team names. A handful of countries
come through with a stray `rn">` stuck to the front, which is a leftover
from however the data was scraped. I strip that out, trim any spare
spaces, and while I am at it I add one new column for the total goals
scored in each game.

```{r clean2}
matches <- matches %>%
  mutate(
    `Home Team Name` = str_remove_all(`Home Team Name`, 'rn">') %>% str_trim(),
    `Away Team Name` = str_remove_all(`Away Team Name`, 'rn">') %>% str_trim(),
    total_goals = `Home Team Goals` + `Away Team Goals`
  )

glimpse(matches %>% select(Year, Stage, `Home Team Name`, `Away Team Name`,
                           `Home Team Goals`, `Away Team Goals`, total_goals,
                           Attendance))
```

# A quick summary of the tournaments

Before plotting anything I like to look at the numbers in a plain table.
Here is one row per tournament with the number of games, the average
goals per game, and the average crowd.

```{r summary-table}
by_year <- matches %>%
  group_by(Year) %>%
  summarize(
    games = n(),
    goals_per_game = round(mean(total_goals), 2),
    avg_attendance = round(mean(Attendance, na.rm = TRUE))
  ) %>%
  ungroup()

by_year
```

You can already see the shape of two stories in that table. The
tournament keeps getting bigger, from eighteen games in 1930 to sixty
four by the late 1990s, and the goals per game were much higher in the
early years than they are now. The charts below make those patterns
easier to feel.

# Chart 1: Goals per game have cooled off since the 1950s

My first question was simple. Are World Cup games as full of goals as
they used to be? I plot the average goals per game for each tournament
and mark the highest and lowest points so the eye lands on them straight
away.

```{r chart-goals}
peak <- by_year %>% slice_max(goals_per_game, n = 1)
low  <- by_year %>% slice_min(goals_per_game, n = 1)

ggplot(by_year, aes(x = Year, y = goals_per_game)) +
  geom_line(color = "#9aa7b1", linewidth = 1) +
  geom_point(color = "#2c7fb8", size = 2.6) +
  geom_point(data = peak, color = "#e8702a", size = 3.6) +
  geom_point(data = low,  color = "#e8702a", size = 3.6) +
  geom_text(data = peak,
            aes(label = paste0(Year, ": ", goals_per_game)),
            vjust = -1.1, color = "#e8702a", fontface = "bold", size = 3.4) +
  geom_text(data = low,
            aes(label = paste0(Year, ": ", goals_per_game)),
            vjust = 1.9, color = "#e8702a", fontface = "bold", size = 3.4) +
  scale_x_continuous(breaks = by_year$Year) +
  scale_y_continuous(limits = c(0, 6)) +
  labs(
    title = "Goals per game have cooled off since the 1950s",
    subtitle = "Average goals scored in a World Cup match, by tournament",
    x = NULL,
    y = "Goals per game"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The 1954 tournament in Switzerland was wild, with well over five goals a
game on average. From there the trend drifts down and bottoms out in
1990, the famous tournament that pushed the sport to change some of its
rules afterward. Modern World Cups have settled into a fairly steady
rhythm of a little under three goals a game.

# Chart 2: The crowds keep growing

My second question was about the people in the stands rather than the
players on the pitch. Has the World Cup grown into the giant event it is
talked about as today? I show the average crowd per game for each
tournament.

```{r chart-attendance}
ggplot(by_year, aes(x = factor(Year), y = avg_attendance)) +
  geom_col(fill = "#2c7fb8") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The World Cup grew into a global spectacle?",
    subtitle = "Average attendance per match, by tournament",
    x = NULL,
    y = "Average attendance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The crowds are not a perfectly smooth climb, because the size of the
host country and its stadiums matters a lot. The huge 1994 average
belongs to the United States, where games were played in enormous
American football stadiums. Still, the overall direction is clear. A
typical World Cup match now draws roughly twice the crowd it did in the
early decades.

# Chart 3: The countries that score the most

My last question is the one every fan argues about. Which countries have
put the ball in the net the most across World Cup history? Because the
data splits goals into a home side and an away side for each game, I add
up both for every country and keep the top ten.

```{r chart-teams}
home_goals <- matches %>%
  group_by(team = `Home Team Name`) %>%
  summarize(goals = sum(`Home Team Goals`), .groups = "drop")

away_goals <- matches %>%
  group_by(team = `Away Team Name`) %>%
  summarize(goals = sum(`Away Team Goals`), .groups = "drop")

top_teams <- bind_rows(home_goals, away_goals) %>%
  group_by(team) %>%
  summarize(goals = sum(goals), .groups = "drop") %>%
  slice_max(goals, n = 10) %>%
  mutate(is_brazil = team == "Brazil")

top_teams

ggplot(top_teams, aes(x = goals, y = reorder(team, goals), fill = is_brazil)) +
  geom_col() +
  geom_text(aes(label = goals), hjust = -0.2, size = 3.4) +
  scale_fill_manual(values = c(`TRUE` = "#e8702a", `FALSE` = "#2c7fb8")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Brazil sits on top of the scoring chart",
    subtitle = "Total goals scored across all World Cup matches, 1930 to 2014",
    x = "Total goals",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "none"
  )
```

Brazil is way out in front, which fits its reputation as the most
successful country in the tournament. After that the usual European and
South American powers fill out the list. The split between Germany and
Germany FR is worth a note, because the dataset keeps the old West
Germany record separate from the reunified Germany, so the country is
really even stronger than its single bar suggests.

# Wrapping up the analysis

Three plain questions, three plain charts. Goals per game have come down
from their wild peak in the fifties. The crowds have grown as the
tournament turned into a worldwide event. And a small group of
countries, led by Brazil, has done most of the scoring.

------------------------------------------------------------------------

# Project write up

## What charts did I plan at the start?

My very first idea was actually a lot busier than what I ended up with.
I wanted to plot every single match as a dot on a timeline, colored by
how many goals were scored, almost like a scatter of the whole history
at once. I also thought about a map of the host cities and a chart
breaking goals down by the stage of the tournament, like group games
versus finals.

Once I started sketching those, I realized they were either too
cluttered to read or needed data the file does not really hold in a
clean way. The match by match scatter was a wall of dots with no clear
message. So I stepped back and asked what a normal fan would actually
want to know, and that gave me the three questions I built the report
around: are there fewer goals now, are the crowds bigger, and who scores
the most.

## What story do the charts tell?

Put together, the charts tell the story of a sport that got bigger and
tighter at the same time. The early World Cups were small affairs with
smaller crowds but buckets of goals. As the tournament grew into a
global event with packed stadiums, the football itself became more
careful and defensive, and the goals per game settled down to a steadier
number. The last chart adds the human side of that story, which is that
a handful of footballing nations, with Brazil at the front, have written
most of the history.

## How did I apply the ideas of good data visualization?

I tried to keep a few principles from the course in mind the whole way
through.

I led with the message, not the chart type. Every title is a full
sentence that states the takeaway, so a reader who only looks at the
headings still gets the point. The subtitles then explain exactly what
the axes are measuring.

I removed clutter. I used a clean minimal theme, dropped the heavy grid
lines I did not need, and let white space do some of the work. Nothing
on the page is there for decoration alone.

I used color with a purpose. The charts mostly stick to one calm blue,
and I save a single warm orange for the things I want to notice, like
the highest and lowest scoring years and Brazil's bar. Because color is
rare, it actually means something when it shows up.

I made the numbers easy to compare. The team chart is sorted from most
goals to fewest and turned on its side so the country names are easy to
read, and I put the exact totals at the end of each bar so you do not
have to guess. The attendance numbers are formatted with commas so a big
crowd reads as a big crowd at a glance.

Finally, I respected the data. I cleaned out the blank rows, the
duplicates, and the broken team names before plotting anything, because
a pretty chart built on messy numbers is worse than no chart at all.
