---
title: "A Century of World Cup Football"
author: "Niklas Grau (ngrau3035@floridapoly.edu)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    df_print: kable
    fig_width: 8
    fig_height: 5
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(tidyverse)
library(scales)
library(plotly)   # for the interactive chart further down
```

# Introduction

This little project looks at every match played at the FIFA World Cup
from the first tournament in Uruguay in 1930 up to Brazil in 2014. The
file `WorldCupMatches.csv` lists one row per game and records who
played, how many goals each side scored, where the game took place, and
how many people came to watch.

I grew up watching the World Cup with my family, so when I saw this
dataset in the list I knew right away it was the one I wanted to dig
into. My goal was not to build something fancy. I wanted three clean
charts that each answer one simple question a casual fan might actually
ask.

# Loading the data

I keep the raw download in the `data/` folder and read it from there so
the whole report runs the same way on any machine.

```{r load}
matches_raw <- read_csv("../data/WorldCupMatches.csv", show_col_types = FALSE)

dim(matches_raw)
```

# Getting the data ready

The first thing I noticed is that the raw file is messy. It has
thousands of completely blank rows hanging off the bottom, and many of
the real games are repeated. Before doing anything else I drop the empty
rows and remove exact duplicates.

```{r clean1}
matches <- matches_raw %>%
  filter(!is.na(MatchID)) %>%   # blank rows have no match id
  distinct()                    # drop repeated rows

dim(matches)
```

That takes the file from over four thousand rows down to the real set of
games.

A second problem hides inside the team names. A handful of countries
come through with a stray `rn">` stuck to the front, which is a leftover
from however the data was scraped. I strip that out, trim any spare
spaces, and while I am at it I add one new column for the total goals
scored in each game.

```{r clean2}
matches <- matches %>%
  mutate(
    `Home Team Name` = str_remove_all(`Home Team Name`, 'rn">') %>% str_trim(),
    `Away Team Name` = str_remove_all(`Away Team Name`, 'rn">') %>% str_trim(),
    total_goals = `Home Team Goals` + `Away Team Goals`
  )

glimpse(matches %>% select(Year, Stage, `Home Team Name`, `Away Team Name`,
                           `Home Team Goals`, `Away Team Goals`, total_goals,
                           Attendance))
```

# A quick summary of the tournaments

Before plotting anything I like to look at the numbers in a plain table.
Here is one row per tournament with the number of games, the average
goals per game, and the average crowd.

```{r summary-table}
by_year <- matches %>%
  group_by(Year) %>%
  summarize(
    games = n(),
    goals_per_game = round(mean(total_goals), 2),
    avg_attendance = round(mean(Attendance, na.rm = TRUE))
  ) %>%
  ungroup()

by_year
```

You can already see the shape of two stories in that table. The
tournament keeps getting bigger, from eighteen games in 1930 to sixty
four by the late 1990s, and the goals per game were much higher in the
early years than they are now. The charts below make those patterns
easier to feel.

# Picking colors everyone can read

Before I draw anything I want to settle on colors that work for readers
who are colorblind, since red and green together are exactly the pair a
lot of people struggle to tell apart. I use the Okabe and Ito palette
here, which was built to stay readable under every common form of color
blindness. One blue for the normal data, one strong orange for the thing
I want you to notice, and a plain grey for background lines.

```{r palette}
main_blue <- "#0072B2"   
accent    <- "#D55E00"   
context   <- "#9aa7b1"   
```

# Chart 1: Goals per game have cooled off since the 1950s

My first question was simple. Are World Cup games as full of goals as
they used to be? I plot the average goals per game for each tournament
and mark the highest and lowest points so the eye lands on them straight
away.

```{r chart-goals, fig.alt="Line chart of the average goals scored per World Cup match for each tournament from 1930 to 2014. The line climbs to a peak of about 5.4 goals per game in 1954, then drifts downward over the decades to a low of about 2.2 in 1990, and settles a little under three goals per game in the most recent tournaments. The highest and lowest years are highlighted in orange and labeled."}
peak <- by_year %>% slice_max(goals_per_game, n = 1)
low  <- by_year %>% slice_min(goals_per_game, n = 1)

ggplot(by_year, aes(x = Year, y = goals_per_game)) +
  geom_line(color = context, linewidth = 1) +
  geom_point(color = main_blue, size = 2.6) +
  geom_point(data = peak, color = accent, size = 3.6) +
  geom_point(data = low,  color = accent, size = 3.6) +
  geom_text(data = peak,
            aes(label = paste0(Year, ": ", goals_per_game)),
            vjust = -1.1, color = accent, fontface = "bold", size = 3.4) +
  geom_text(data = low,
            aes(label = paste0(Year, ": ", goals_per_game)),
            vjust = 1.9, color = accent, fontface = "bold", size = 3.4) +
  scale_x_continuous(breaks = by_year$Year) +
  scale_y_continuous(limits = c(0, 6)) +
  labs(
    title = "Goals per game have cooled off since the 1950s",
    subtitle = "Average goals scored in a World Cup match, by tournament",
    x = NULL,
    y = "Goals per game"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The 1954 tournament in Switzerland was wild, with well over five goals a
game on average. From there the trend drifts down and bottoms out in
1990, the famous tournament that pushed the sport to change some of its
rules afterward. Modern World Cups have settled into a fairly steady
rhythm of a little under three goals a game.

# Chart 2: The crowds keep growing

My second question was about the people in the stands rather than the
players on the pitch. Has the World Cup grown into the giant event it is
talked about as today? I show the average crowd per game for each
tournament.

```{r chart-attendance, fig.alt="Bar chart of the average attendance per match at each World Cup from 1930 to 2014. Bars start at zero. Early tournaments average roughly twenty to thirty thousand fans per game, while recent tournaments sit around fifty thousand. The tallest bar is 1994, hosted by the United States, at close to seventy thousand."}
ggplot(by_year, aes(x = factor(Year), y = avg_attendance)) +
  geom_col(fill = main_blue) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The crowds kept growing as the tournament went global",
    subtitle = "Average attendance per match, by tournament",
    x = NULL,
    y = "Average attendance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The crowds are not a perfectly smooth climb, because the size of the
host country and its stadiums matters a lot. The huge 1994 average
belongs to the United States, where games were played in enormous
American football stadiums. Still, the overall direction is clear. A
typical World Cup match now draws roughly twice the crowd it did in the
early decades.

# Chart 3: The countries that score the most

My last question is the one every fan argues about. Which countries have
put the ball in the net the most across World Cup history? Because the
data splits goals into a home side and an away side for each game, I add
up both for every country and keep the top ten.

```{r chart-teams, fig.alt="Horizontal bar chart of the ten countries with the most total goals across all World Cup matches from 1930 to 2014. Bars are sorted from most to fewest goals and each bar is labeled with its exact total. Brazil leads clearly with 221 goals and is highlighted in orange, followed by Argentina, Germany FR, Italy, and France."}
home_goals <- matches %>%
  group_by(team = `Home Team Name`) %>%
  summarize(goals = sum(`Home Team Goals`), .groups = "drop")

away_goals <- matches %>%
  group_by(team = `Away Team Name`) %>%
  summarize(goals = sum(`Away Team Goals`), .groups = "drop")

top_teams <- bind_rows(home_goals, away_goals) %>%
  group_by(team) %>%
  summarize(goals = sum(goals), .groups = "drop") %>%
  slice_max(goals, n = 10) %>%
  mutate(is_brazil = team == "Brazil")

top_teams

ggplot(top_teams, aes(x = goals, y = reorder(team, goals), fill = is_brazil)) +
  geom_col() +
  geom_text(aes(label = goals), hjust = -0.2, size = 3.4) +
  scale_fill_manual(values = c(`TRUE` = accent, `FALSE` = main_blue)) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Brazil sits on top of the scoring chart",
    subtitle = "Total goals scored across all World Cup matches, 1930 to 2014",
    x = "Total goals",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "none"
  )
```

Brazil is way out in front, which fits its reputation as the most
successful country in the tournament. After that the usual European and
South American powers fill out the list. The split between Germany and
Germany FR is worth a note, because the dataset keeps the old West
Germany record separate from the reunified Germany, so the country is
really even stronger than its single bar suggests.

# An interactive look at every tournament

The three charts above each answer one question on their own, but the
numbers behind them all live in the same little per tournament table. So
for the interactive piece I wanted one chart where you can pull up the
whole record of any single World Cup just by pointing at it. The static
charts force me to split goals and crowds across two pictures. Here they
sit together, and the reader decides which tournament to inspect.

Each dot is one tournament. Its position left to right is the year, its
height is the goals per game, the size of the dot is how many games were
played, and the color is the average crowd. Hovering over a dot brings
up the exact figures for that year.

```{r interactive, fig.alt="Interactive scatter plot of World Cup tournaments from 1930 to 2014. Each point is one tournament, positioned by year and by average goals per game, sized by the number of games played, and colored by average attendance. Hovering reveals the exact goals per game, game count, and average crowd for that year."}
plot_df <- by_year %>%
  mutate(
    hover = paste0(
      "Year: ", Year,
      "<br>Goals per game: ", goals_per_game,
      "<br>Games played: ", games,
      "<br>Avg attendance: ", comma(avg_attendance)
    )
  )

gg_int <- ggplot(plot_df, aes(x = Year, y = goals_per_game, text = hover)) +
  geom_line(color = context, linewidth = 0.7) +
  geom_point(aes(size = games, color = avg_attendance)) +
  scale_color_viridis_c(labels = comma, name = "Avg crowd") +
  scale_size_continuous(name = "Games", range = c(2, 7)) +
  labs(
    title = "Every World Cup, in one view",
    subtitle = "Point for each tournament. Hover for the full record.",
    x = NULL,
    y = "Goals per game"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())

ggplotly(gg_int, tooltip = "text")
```

The viridis color scale I use here is one of those palettes that stays
readable for colorblind viewers, so the crowd story comes through for
everyone and not just for people who see the full range of colors.

# Fixing a bad chart: before and after

The assignment asked me to take one chart and improve it, so I went back
to the attendance chart from earlier. It does its job, but it leaves one
trap open for the reader, so this is the chart I picked to fix.

## The before

```{r before, fig.alt="The original attendance bar chart, all bars in the same blue. The 1994 bar is the tallest at close to seventy thousand, but nothing on the chart points it out or explains why, so a reader could mistake it for a sudden jump in popularity."}
ggplot(by_year, aes(x = factor(Year), y = avg_attendance)) +
  geom_col(fill = main_blue) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The crowds kept growing as the tournament went global",
    subtitle = "Average attendance per match, by tournament",
    x = NULL,
    y = "Average attendance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

The chart itself is honest, the axis starts at zero and the bars are the
right height, but it still leaves a trap open. The 1994 bar is the
tallest on the whole chart, and with nothing pointing it out a reader
could easily walk away thinking the World Cup suddenly got far more
popular that year. The real reason is a lot more boring. The 1994
tournament was hosted in the United States and many of the games were
played in enormous American football stadiums, so the average crowd was
pushed up by the size of the venues rather than by any jump in interest.
The title is fine as a general headline, but it only names the overall
trend and never warns the reader about the one bar that is most likely
to be misread.

## The after

```{r after, fig.alt="The improved version of the same chart. The 1994 bar is now colored orange and a short note above it explains that the spike comes from the huge US football stadiums. Every other bar stays one blue, the axis still starts at zero, the numbers use commas, and the title now states the takeaway."}
att_df <- by_year %>% mutate(is_1994 = Year == 1994)

ggplot(att_df, aes(x = factor(Year), y = avg_attendance, fill = is_1994)) +
  geom_col() +
  annotate("text", x = "1994", y = 71500,
           label = "1994: huge US\nfootball stadiums",
           size = 3, color = accent, fontface = "bold", lineheight = 0.9) +
  scale_fill_manual(values = c(`TRUE` = accent, `FALSE` = main_blue)) +
  scale_y_continuous(labels = comma, limits = c(0, 78000)) +
  labs(
    title = "A typical World Cup match now draws about fifty thousand fans",
    subtitle = "Average attendance per match, by tournament",
    x = NULL,
    y = "Average attendance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

The improved version fixes that. The single bar for 1994 is now colored
orange so the eye lands on it right away, and a short note sits just
above it explaining that the spike comes from the huge US stadiums. That
way the one bar that actually needs context gets it, instead of quietly
misleading anyone who only skims the chart.

# Wrapping up the analysis

Three plain questions, three plain charts. Goals per game have come down
from their wild peak in the fifties. The crowds have grown as the
tournament turned into a worldwide event. And a small group of
countries, led by Brazil, has done most of the scoring.

------------------------------------------------------------------------

# Project write up

## What charts did I plan at the start?

My very first idea was actually a lot busier than what I ended up with.
I wanted to plot every single match as a dot on a timeline, colored by
how many goals were scored, almost like a scatter of the whole history
at once. I also thought about a map of the host cities and a chart
breaking goals down by the stage of the tournament, like group games
versus finals.

Once I started sketching those, I realized they were either too
cluttered to read or needed data the file does not really hold in a
clean way. The match by match scatter was a wall of dots with no clear
message. So I stepped back and asked what a normal fan would actually
want to know, and that gave me the three questions I built the report
around: are there fewer goals now, are the crowds bigger, and who scores
the most.

## What story do the charts tell?

Put together, the charts tell the story of a sport that got bigger and
tighter at the same time. The early World Cups were small affairs with
smaller crowds but buckets of goals. As the tournament grew into a
global event with packed stadiums, the football itself became more
careful and defensive, and the goals per game settled down to a steadier
number. The last chart adds the human side of that story, which is that
a handful of footballing nations, with Brazil at the front, have written
most of the history.

## How did I apply the ideas of good data visualization?

I tried to keep a few principles from the course in mind the whole way
through.

I led with the message, not the chart type. Every title is a full
sentence that states the takeaway, so a reader who only looks at the
headings still gets the point. The subtitles then explain exactly what
the axes are measuring.

I removed clutter. I used a clean minimal theme, dropped the heavy grid
lines I did not need, and let white space do some of the work. Nothing
on the page is there for decoration alone.

I used color with a purpose. The charts mostly stick to one calm blue,
and I save a single warm orange for the things I want to notice, like
the highest and lowest scoring years and Brazil's bar. Because color is
rare, it actually means something when it shows up.

I made the numbers easy to compare. The team chart is sorted from most
goals to fewest and turned on its side so the country names are easy to
read, and I put the exact totals at the end of each bar so you do not
have to guess. The attendance numbers are formatted with commas so a big
crowd reads as a big crowd at a glance.

Finally, I respected the data. I cleaned out the blank rows, the
duplicates, and the broken team names before plotting anything, because
a pretty chart built on messy numbers is worse than no chart at all.

## Accessibility, interactivity, and the redesign

I also tried to make the report work for as many readers as possible.
Every figure has alt text written into the chunk so a screen reader can
describe it, the colors come from palettes that stay readable for
colorblind viewers, and wherever color carries a message I backed it up
with a label or a sorted order so nobody has to rely on color alone.

The interactive tournament chart is the part I had the most fun with. A
static chart shows you one fixed view, but this one lets the reader
hover to read the exact goals, games, and crowd for any World Cup, and
zoom into the busy modern years instead of squinting at a cluster. It
turns the per tournament table into something you can actually poke at.
