---
title: "What Makes an Elite Footballer? A Look at FIFA 18 Player Data"
author: "Niklas Grau - ngrau3035@floridapoly.edu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Global chunk options. I want the code visible but the messages and
# warnings hidden so the report reads cleanly.
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5.5,
  dpi = 120
)
```

## Introduction

I started playing FIFA when I was a kid, thats why I chose this dataset.
The FIFA 18 dataset is one of those rare sports files that is both huge
and tidy. EA Sports assigns every player in the game a long list of
numeric attributes, from sprint speed to composure, and rolls all of
that into a single headline number called the overall rating.

The thread running through this report is a single question: what
separates the truly elite players from everyone else, and where in the
world do those players tend to come from? Along the way I look at how
ratings change as players age, how much room younger players still have
to grow, which countries produce the strongest squads on average, and
finally which on-pitch attributes the rating actually rewards the most.

## Packages

```{r packages}
library(tidyverse)   # dplyr, ggplot2, readr, tidyr, forcats, stringr
library(plotly)      # the interactive scatter plot
library(sf)          # reading the shapefile and drawing the map
library(broom)       # turning the linear model into a tidy data frame
library(scales)      # nicer axis labels
library(htmlwidgets) # saving the interactive plot as a standalone HTML file
```

## A consistent look

Before plotting anything I set up a small palette and a custom theme so
that every figure shares the same colors, fonts, and spacing. Keeping
this in one place means I can change the look of the whole report by
editing a single block.

```{r theme}
# A simple football themed palette. Deep pitch green as the main color,
# a warm orange for callouts and highlights, and a soft grey for context.
pitch_green  <- "#0b6e4f"
warm_accent  <- "#e0701b"
soft_grey    <- "#9aa0a6"

theme_pitch <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "grey30", margin = margin(b = 10)),
      plot.caption  = element_text(color = "grey50", size = 9),
      panel.grid.minor = element_blank(),
      axis.title    = element_text(color = "grey25"),
      legend.position = "right"
    )
}

# Make the theme the default for every ggplot in the document.
theme_set(theme_pitch())
```

With the theme set as the default, I do not have to remember to add it
to every plot.

## Loading and inspecting the data

```{r load}
players <- read_csv("../data/fifa18.csv", show_col_types = FALSE)

dim(players)
```

The file holds 17076 players described by 40 columns. The first few
columns are descriptive (name, nationality, club, age) and the rest are
the skill attributes that the game tracks.

```{r}
players
```

```{r glimpse}
glimpse(players[, 1:12])
```

```{r missing}
# Quick scan for missing values across every column.
players %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  filter(n_missing > 0)
```

The missing value check comes back empty, which is a pleasant surprise.
There is nothing to impute or drop, so I can move straight on to the
analysis. The numeric attributes all sit on a familiar zero to 100
scale, and a fast look at the ranges confirms that age runs from the mid
teens up to the late forties and overall ratings span from the mid
forties up to ninety four.

```{r ranges}
players %>%
  summarise(
    age_min     = min(age),     age_max     = max(age),
    overall_min = min(overall), overall_max = max(overall),
    n_countries = n_distinct(nationality),
    n_clubs     = n_distinct(club)
  )
```

So we have a global pool: more than one hundred and sixty nationalities
spread across several hundred clubs. That spread is exactly what makes
the map later worth drawing.

## Visualization 1: the career arc, with annotations

My first question is the most intuitive one. How does a player's overall
rating change as they get older? Footballers are supposed to peak
somewhere in their late twenties, so I want to see whether the game's
ratings agree.

```{r age-summary}
# Average overall rating at each age, keeping only ages with a
# reasonable number of players so the tails are not noisy.
age_curve <- players %>%
  group_by(age) %>%
  summarise(mean_overall = mean(overall), n = n(), .groups = "drop") %>%
  filter(n >= 15)

# Find the age where the average rating peaks so I can label it.
peak <- age_curve %>% slice_max(mean_overall, n = 1)
peak
```

```{r age-plot, fig.height=5.5, fig.alt="Line and point chart of the average FIFA 18 overall rating at each age, for ages that have at least fifteen players. Ratings climb through the early twenties, plateau, and peak around age 31 before slowly declining. Each point is sized by how many players are at that age, showing most players are in their early and middle twenties. A dashed vertical line and a text note mark the peak age."}
ggplot(age_curve, aes(x = age, y = mean_overall)) +
  geom_line(color = soft_grey, linewidth = 0.8) +
  geom_point(aes(size = n), color = pitch_green, alpha = 0.8) +
  # A vertical guide and a text callout marking the peak age.
  geom_vline(xintercept = peak$age, linetype = "dashed", color = warm_accent) +
  annotate(
    "text",
    x = peak$age + 0.5, y = min(age_curve$mean_overall) + 1,
    label = paste0("Ratings peak around age ", peak$age),
    hjust = 0, color = warm_accent, fontface = "bold"
  ) +
  annotate(
    "text",
    x = 17, y = max(age_curve$mean_overall) - 0.5,
    label = "Teenagers are rated low\non current ability",
    hjust = 0, color = "grey40", size = 3.4, lineheight = 0.95
  ) +
  scale_size_continuous(name = "Players at this age", range = c(1, 6), labels = comma) +
  labs(
    title = "A footballer's rating climbs into the late twenties, then eases down",
    subtitle = "Average FIFA 18 overall rating by age, for ages with at least 15 players",
    x = "Age",
    y = "Average overall rating"
  )
```

The curve tells a continuous story. Average ratings climb steadily
through a player's early twenties, flatten out into a broad plateau, and
reach their high point around age of 31. After that the line drifts
down, but slowly, which fits the idea that experience and composure keep
older players useful even as their physical attributes fade. The dot
sizes are worth noticing too, since most players in the game are
clustered in their early and middle twenties, and the very high and very
low ages are thin on numbers.

## Visualization 2: room to grow (interactive)

The age curve hints that young players are rated low because the game
scores their current ability, not their ceiling. FIFA 18 actually tracks
that ceiling directly with a separate potential rating. Plotting current
overall against potential should show the gap, and an interactive chart
lets the reader hover over individual players to see who sits where.

```{r interactive-data}
# Focus on players rated 70 or higher. That keeps the chart responsive
# and centers it on players who actually matter competitively.
talent <- players %>%
  filter(overall >= 70) %>%
  mutate(
    growth = potential - overall,
    # Build a tidy hover label for each point.
    label = paste0(
      name, "<br>",
      club, " (", nationality, ")<br>",
      "Age: ", age, "<br>",
      "Overall: ", overall, "  Potential: ", potential
    )
  )

nrow(talent)
```

```{r interactive-plot, fig.alt="Interactive scatter plot of FIFA 18 players rated 70 or higher, plotting current overall rating against potential rating. A dashed diagonal marks where potential equals current ability. Points are colored by age using a colorblind-safe viridis scale, and hovering reveals each player's name, club, nationality, age, overall, and potential. Younger players sit above the line with room to grow while older players hug it."}
gg_growth <- ggplot(talent, aes(x = overall, y = potential, text = label)) +
  # The reference line where potential equals current ability.
  geom_abline(slope = 1, intercept = 0, color = soft_grey, linetype = "dashed") +
  geom_point(aes(color = age), alpha = 0.6, size = 1.6) +
  scale_color_viridis_c(name = "Age") +
  labs(
    title = "How much room does a player still have to grow?",
    subtitle = "Players rated 70 or above. Points above the line still have upside.",
    x = "Current overall rating",
    y = "Potential rating"
  )

# Turn the ggplot into an interactive plotly chart driven by the hover text.
interactive_growth <- ggplotly(gg_growth, tooltip = "text")
interactive_growth
```

```{r save-widget}
# Save a fully self contained copy of the interactive chart so it can be
# opened on its own without the rest of the report.
saveWidget(
  interactive_growth,
  file = "interactive_potential_overall.html",
  selfcontained = TRUE
)
```

Reading this chart is mostly about the dashed diagonal. Every point on
the line is a player whose potential matches their current rating, which
is the game's way of saying they are as good as they are going to get.
Points that float above the line still have upside, and the color makes
the pattern obvious, since the darker younger players sit high above the
line while the brighter older players hug it. This is the chart where
being interactive really earns its keep. A static picture of seventeen
hundred overlapping dots would be a blob, but here you can hover any
point to pull up the player's name, club, age, and both ratings.

## Visualization 3: where the talent lives (spatial)

If individual players have a story, so countries do as well. I want a
world map shaded by the average overall rating of each nation's players.
This needs two pieces: a summary of the ratings by country, and a
shapefile that knows the shape of each country.

The tricky part is that the names do not line up cleanly. FIFA lists
England, Scotland, Wales, and Northern Ireland separately, while the map
only knows the United Kingdom. Several other countries use different
spellings between the two sources. I handle this with an explicit recode
step, and I fold the four home nations into the United Kingdom so that
England's large player pool does not simply vanish from the map.

```{r country-recode}
# Translate FIFA nationality names into the names used by the shapefile.
# Anything not listed here already matches, so it passes through unchanged.
recode_country <- function(x) {
  dplyr::recode(x,
    "England"              = "United Kingdom",
    "Scotland"             = "United Kingdom",
    "Wales"                = "United Kingdom",
    "Northern Ireland"     = "United Kingdom",
    "Republic of Ireland"  = "Ireland",
    "United States"        = "United States of America",
    "Korea Republic"       = "South Korea",
    "Korea DPR"            = "North Korea",
    "Serbia"               = "Republic of Serbia",
    "Czech Republic"       = "Czechia",
    "DR Congo"             = "Democratic Republic of the Congo",
    "Congo"                = "Republic of the Congo",
    "Bosnia Herzegovina"   = "Bosnia and Herzegovina",
    "FYR Macedonia"        = "North Macedonia",
    "Guinea Bissau"        = "Guinea-Bissau",
    "Central African Rep." = "Central African Republic",
    "China PR"             = "China",
    "Trinidad & Tobago"    = "Trinidad and Tobago",
    .default = x
  )
}

country_stats <- players %>%
  mutate(country = recode_country(nationality)) %>%
  group_by(country) %>%
  summarise(mean_overall = mean(overall), n_players = n(), .groups = "drop") %>%
  # Require a minimum squad size so a single star cannot dominate a country.
  filter(n_players >= 20)

# The strongest countries by average rating, just to sanity check the result.
country_stats %>% slice_max(mean_overall, n = 8)
```

```{r read-shapefile}
# Read the Natural Earth country outlines and drop Antarctica.
world <- st_read("../data/ne_countries/ne_110m_admin_0_countries.shp", quiet = TRUE) %>%
  filter(ADMIN != "Antarctica")

# Attach the rating summary to each country polygon.
world_rated <- world %>%
  left_join(country_stats, by = c("ADMIN" = "country"))
```

```{r map-plot, fig.height=5.5, fig.alt="World map shaded by the average FIFA 18 overall rating of each country's players, for countries with at least twenty players. A colorblind-safe viridis scale runs from light for lower averages to dark for higher averages. Countries with too few players are left grey. Strength is spread fairly widely across Europe and South America with pockets elsewhere."}
# A label for the single highest rated country, placed at its centroid.
top_country <- country_stats %>% slice_max(mean_overall, n = 1)

ggplot(world_rated) +
  geom_sf(aes(fill = mean_overall), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    option = "mako", direction = -1,
    na.value = "grey85", name = "Average\noverall"
  ) +
  coord_sf(crs = "+proj=robin") +
  labs(
    title = "Average player rating by country",
    subtitle = paste0(
      "Countries with at least 20 players. ",
      top_country$country, " has the highest average."
    ),
    caption = "Grey countries had fewer than 20 players in the data."
  ) +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

The darkest countries are not always the ones you would expect, because
this is an average rather than a count, so a small footballing nation
that only sends its very best players abroad can score higher than a
country that produces thousands of players at every level. The countries
shaded grey simply did not clear the twenty player threshold, which is
why a lot of the map is blank. The headline is that strength in this
game is spread fairly widely across Europe and South America, with
pockets elsewhere, rather than being concentrated in one corner of the
world.

## Visualization 4: what the rating actually rewards (model)

The last question is the most analytical. The overall rating is a single
number, but it is built from dozens of attributes. Which of those
attributes carry the most weight? A linear model is a clean way to ask
this. I regress overall rating on a handful of skill attributes, and
because the attributes live on the same zero to 100 scale but have
different spreads, I standardize them first so the coefficients can be
compared directly.

```{r model-fit}
# Pick a set of attributes that cover passing, finishing, physicality,
# and mentality, then standardize each one to mean zero and unit variance.
model_data <- players %>%
  select(overall, reactions, composure, short_passing, vision,
         ball_control, strength, stamina, interceptions,
         finishing, sprint_speed) %>%
  mutate(across(-overall, ~ as.numeric(scale(.))))

fit <- lm(overall ~ ., data = model_data)

# Pull out the model fit summary.
glance(fit) %>% select(r.squared, adj.r.squared, sigma)
```

```{r model-coef-plot, fig.height=5.5, fig.alt="Coefficient plot from a linear model of overall rating on eleven standardized player attributes, sorted by effect size. Reactions has by far the largest positive effect, followed by composure, ball control, and short passing. A few attributes such as finishing and interceptions show small negative coefficients. Each estimate is drawn with a 95 percent confidence interval, all of them tight because the sample is large."}
# Tidy the coefficients with confidence intervals, drop the intercept,
# and tidy up the attribute names for the labels.
coefs <- tidy(fit, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = str_replace_all(term, "_", " ")) %>%
  arrange(estimate)

ggplot(coefs, aes(x = estimate, y = fct_reorder(term, estimate))) +
  geom_vline(xintercept = 0, color = soft_grey, linetype = "dashed") +
  geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high),
    color = pitch_green, fatten = 3
  ) +
  labs(
    title = "Which attributes drive a player's overall rating?",
    subtitle = "Standardized coefficients from a linear model.",
    x = "Effect on overall rating (per one standard deviation)",
    y = NULL,
    caption = "Horizontal lines show 95 percent confidence intervals."
  )
```

The model explains the rating extremely well, with an R squared above
0.75, which tells us these eleven attributes capture most of what goes
into the headline number. The coefficient plot then ranks them.
Reactions sits far out in front, followed by composure, which is
interesting because neither one is a flashy attribute that a casual fan
like me would name first. The game seems to treat a player's mental
sharpness and ability to stay calm as the backbone of quality, while
crowd pleasing traits like finishing and raw sprint speed matter, but
matter less. The confidence intervals are tight because the sample is so
large, so the ordering here is something we can trust rather than an
artifact of noise.

I am not sure why we get negative coefficients for attributes like
finishing and interceptions. My assumption is, that these are an
artifact of the attributes being highly correlated and of pooling every
position into one model, so they should not be taken at face value.

## Fixing a bad chart: before and after

The assignment asked me to take one chart and improve it. I went back to
the age curve from the start of the report, because the first version I
drew of it had a real problem hiding in the tail.

### The before

```{r before-fifa, fig.height=5.5, fig.alt="The naive age curve, plotting average overall rating for every age in the data with no minimum sample size. The line is smooth through the twenties and thirties but then whips up and down in the forties, dropping to around 55, jumping past 70, and crashing again, which makes it look like players fall apart and then suddenly turn elite late in their careers."}
# Average overall at EVERY age, with no minimum count behind each point.
# This is the version that gets the tail badly wrong.
age_naive <- players %>%
  group_by(age) %>%
  summarise(mean_overall = mean(overall), .groups = "drop")

ggplot(age_naive, aes(x = age, y = mean_overall)) +
  geom_line(color = pitch_green, linewidth = 0.8) +
  geom_point(color = pitch_green, size = 1.8) +
  labs(
    title = "A footballer's rating over age",
    subtitle = "Average FIFA 18 overall rating by age, every age included",
    x = "Age",
    y = "Average overall rating"
  )
```

The middle of this chart is fine, but the right tail goes haywire. The
line sinks to around 55 at age 43, leaps past 70 a year later, then
drops off a cliff again. Read literally it says footballers collapse and
then suddenly turn elite in their mid forties, which is plainly not
true. The problem is that those ages are built on almost nobody. Age 44
is two players and age 47 is a single player, yet the chart gives each
of them the same visual weight as age 25, which is averaged over more
than a thousand players. One unusual veteran can swing the whole line,
and nothing on the chart warns the reader that the tail is basically
noise.

### The after

```{r after-fifa, fig.height=5.5, fig.alt="The improved age curve. Ages with fewer than fifteen players are dropped, so the noisy tail is gone and the line becomes a smooth career arc that rises through the twenties, plateaus, and peaks around age 31 before easing down. Each point is sized by the number of players at that age, showing that most players are in their twenties while the older ages are thin. A dashed line and notes mark the peak and the low teenage ratings."}
# Keep only ages backed by a reasonable number of players, and let the
# dot size show where the data is solid.
age_curve <- players %>%
  group_by(age) %>%
  summarise(mean_overall = mean(overall), n = n(), .groups = "drop") %>%
  filter(n >= 15)

peak <- age_curve %>% slice_max(mean_overall, n = 1)

ggplot(age_curve, aes(x = age, y = mean_overall)) +
  geom_line(color = soft_grey, linewidth = 0.8) +
  geom_point(aes(size = n), color = pitch_green, alpha = 0.8) +
  geom_vline(xintercept = peak$age, linetype = "dashed", color = warm_accent) +
  annotate(
    "text",
    x = peak$age + 0.5, y = min(age_curve$mean_overall) + 1,
    label = paste0("Ratings peak around age ", peak$age),
    hjust = 0, color = warm_accent, fontface = "bold"
  ) +
  annotate(
    "text",
    x = 17, y = max(age_curve$mean_overall) - 0.5,
    label = "Teenagers are rated low\non current ability",
    hjust = 0, color = "grey40", size = 3.4, lineheight = 0.95
  ) +
  scale_size_continuous(name = "Players at this age", range = c(1, 6), labels = comma) +
  labs(
    title = "A footballer's rating climbs into the late twenties, then eases down",
    subtitle = "Average FIFA 18 overall rating by age, for ages with at least 15 players",
    x = "Age",
    y = "Average overall rating"
  )
```

The fix is not really about color or style, it is about honesty.
Dropping every age that has fewer than fifteen players clears out the
whole jittery tail in one move, and the line settles into the believable
career arc it should have been all along. Sizing each dot by the number
of players then makes the evidence visible, so the reader can see that
the twenties rest on huge samples while the late thirties are thin. The
dashed line marks where the rating really peaks. Same data and same
question, but the chart now tells the truth instead of letting a few
lonely veterans shout over everyone else. This cleaned up version is the
one I lead with as Visualization 1 at the top of the report.

## Discussion

### What I set out to make, and how the data needed cleaning

My original plan was three charts that built on each other: a
relationship between age and rating, a geographic view of where players
come from, and a model of what drives the rating. The age curve grew an
interactive companion once I noticed the potential column, which gave me
the room to grow chart, and that ended up being the most engaging piece.

The cleaning itself was light in one sense, since there were no missing
values anywhere, but the country names were a genuine headache. The
mismatch between FIFA's home nations and the single United Kingdom
polygon on the map would have quietly dropped more than two thousand
players if I had ignored it. Building the explicit recode function and
folding the home nations together was the single most important
wrangling step in the whole project. The other recurring decision was
filtering. Averages over tiny groups are meaningless, so I set minimum
counts both for the age curve and for the country map.

### The story, the difficulties, and what else I would try

If I had to summarize the report in one sentence, it would be that a
player's value in FIFA 18 is mostly about mentality and arrives in the
late twenties, and that the talent behind it is spread broadly rather
than hoarded by a few countries. Each chart adds a layer to that idea.

The difficulties were mostly practical. Getting seventeen thousand
points into an interactive plot without it becoming sluggish meant
filtering down to the players rated seventy and above, which I think is
a fair trade since those are the players worth inspecting one by one.
The map projection and the name matching took a few tries to get right.
Standardizing the model inputs was important too, because without it the
raw coefficients would have been impossible to compare side by side.

There is plenty more to explore. Splitting players into goalkeepers and
outfield players would sharpen the model, since the goalkeeping
attributes behave very differently from the rest. A count based map
showing where the sheer volume of players comes from would pair nicely
with the average based map already here. And it would be fun to model
potential rather than overall, to see whether the game rewards a
different mix of traits when it is guessing at the future instead of
grading the present.

### How I applied the principles of good visualization

I tried to let each chart make one clear point. The age curve uses a
single line and a couple of annotations so the eye goes straight to the
peak. The interactive scatter leans on a reference line so the reader
has something to measure against, and color carries the secondary age
story without cluttering the axes. The map uses a sequential viridis
scale that runs from light to dark, which matches the natural reading of
low to high, and grey is reserved for missing data so it never competes
with the real values. The model plot is sorted by effect size so the
ranking is the first thing you see, and the confidence intervals are
shown rather than hidden. Throughout, I kept one palette and one theme
so the report feels like a single document rather than four unrelated
images.

I also kept accessibility in mind the whole way. Every figure has alt
text written into its chunk so a screen reader can describe it, and the
continuous color scales on the interactive scatter and the world map now
use viridis, which is built to stay readable for colorblind viewers
rather than relying on a red to green range that many people cannot tell
apart. Where color carries meaning I leaned on something extra too, like
the dashed reference line on the scatter and the sorted order and labels
on the bars, so the message never depends on color alone.

The interactive scatter is the chart that gains the most from being
interactive. The same seventeen hundred points drawn as a flat image
would just overlap into a smudge, but hovering and zooming let the
reader read off any single player and explore the crowded elite corner,
which is something a static picture cannot offer. And the before and
after on the age curve was a good reminder that a chart can be perfectly
honest and still mislead. The first version was not lying, it just let a
handful of forty something players swing the line as hard as ages backed
by thousands. The fix was simply to drop the ages with too little data
behind them and to let the dot sizes show where the evidence is solid.
